Clines from PDE

### 2015-11-21 notes/pde-clines.Rmd
bivariate integral transform

$$
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator{\sgn}{sgn}
\newcommand{\grad}{\nabla}
$$

@@

Suppose that the region includes a type $A$ allele,
and we want the probability that it inherits entirely from the $A$ side.
Let
$$\begin{aligned}
    g(x,A,a,b,\tau) = \P^{x,A}\{ [a,b] \text{ inherits from $A$, $\tau$ generations ago }\} .
\end{aligned}$$
If $W(t)=\{W_1(t),W_2(t),\ldots,W_{n(t)}(t)\}$ is the state of the process $t$ generations ago,
where $W_k=(x_k,z_k,a_k,b_k)$,
with $x_k$ the position and $W_k\in\{A,B\}$ the allele,
then by independence of branches,
$$\begin{aligned}
    g(x,A,a,b,0) &= 1 \\
    g(x,B,a,b,0) &= 0 \\
    g(x,B,a,b,\tau) &= 0 \qquad \text{if } 0 \in [a,b] \\
    g(x,A,a,b,\tau)
    &=
    \E\left[
        \prod_{i=1}^{n(t)} g(x_k,z_k,a_k,b_k,\tau-t)
    \right] \qquad \text{for each } t \le \tau .
\end{aligned}$$
Taking $t = \epsilon$ in the last equation,
so that the probability of branching is $1-e^{-(b-a)\epsilon}=\epsilon(b-a)+O(\epsilon^2)$,
if $0\le a < b$ this implies that
$$\begin{aligned}
\frac{d}{d\tau} 
g(x,A,a,b,\tau)
    &=
    G_{r(a,b)} g(x,A,a,b,\tau)
    +
     \int_a^b g(x,A,a,\theta,\tau)
     \left(
     p(x,\tau)g(x,A,\theta,b,\tau) 
     + (1-p(x,\tau))g(x,B,\theta,b,\tau) 
     \right)d\theta
\end{aligned}$$
where 
$$\begin{aligned}
r(a,b)=\begin{cases}
    \min(|a|,|b|) \qquad &\text{if } ab \ge 0 \\
    0 \qquad &\text{if } 0 \in [a,b] .
\end{cases}
\end{aligned}$$
and $G_r f(x,z)$ is defined above.
On the other hand, if $a<0<b$, then
$$\begin{aligned}
\frac{d}{d\tau} 
g(x,A,a,b,\tau)
    &=
    G_{r(a,b)} g(x,A,a,b,\tau)
    +
     \int_0^b g(x,A,a,\theta,\tau)
     \left(
     p(x,\tau)g(x,A,\theta,b,\tau)
     + (1-p(x,\tau))g(x,B,\theta,b,\tau) 
     \right)d\theta
    +
     \int_a^0 \left(
     p(x,\tau)g(x,A,a,\theta,\tau)
     + (1-p(x,\tau))g(x,B,a,\theta,\tau) 
     \right) g(x,A,\theta,b,\tau) d\theta
\end{aligned}$$

@@

@@

Note that if we think of $u(a,b)$ as a kernel, 
then the map $u(a,b) \mapsto \int_a^b u(a,\theta) u(\theta,b) d\theta$
is similar to the product of the kernel to itself;
on discretization this becomes like a matrix power.
This is implemented in some C code.
```{r solve_branching}
source("unifprod.R")  # for the C part
integral_map <- function (x,eps) {
    # the discretization of the map y(a,b) -> \int_a^b (h(a,u)*h(u,b)-h(a,b)) du
    # x is a square matrix,
    #   eps is the grid spacing between locations
    x2 <- eps*unifprod(x)
}
xgrid <- setup.grid.1D(x.up=-20, x.down=20, N=100)
agrid <- setup.grid.1D(x.up=-0.5, x.down=0.5, N=30)
# we'll save the solution as an array,
#  with dimensions (x,a,b)
# Note we only use the upper triangle a<=b, but oh well.
soln.array <- array(0.0,dim=c(xgrid$N,agrid$N,agrid$N))

@@

>>>>

$$
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\DeclareMathOperator{\sgn}{sgn}
\newcommand{\grad}{\nabla}
$$

@@

Suppose that the region includes a type $A$ allele,
and we want the probability that it inherits entirely from the $A$ side.
Let
$$\begin{aligned}
    g(x,A,a,b,\tau) = \P^{x,A}\{ [a,b] \text{ inherits from $A$, $\tau$ generations ago }\} .
\end{aligned}$$
If $W(t)=\{W_1(t),W_2(t),\ldots,W_{n(t)}(t)\}$ is the state of the process $t$ generations ago,
where $W_k=(x_k,z_k,a_k,b_k)$,
with $x_k$ the position and $W_k\in\{A,B\}$ the allele,
then by independence of branches,
$$\begin{aligned}
    g(x,A,a,b,0) &= 1 \\
    g(x,B,a,b,0) &= 0 \\
    g(x,A,a,b,\tau)
    &=
    \E\left[
        \prod_{i=1}^{n(t)} g(x_k,z_k,a_k,b_k,\tau-t)
    \right] \qquad \text{for each } t \le \tau .
\end{aligned}$$
Taking $t = \epsilon$ in the last equation,
so that the probability of branching is $1-e^{-(b-a)\epsilon}=\epsilon(b-a)+O(\epsilon^2)$,
if $0\le a < b$ this implies that
$$\begin{aligned}
\frac{d}{d\tau} 
g(x,A,a,b,\tau)
    &=
    G_{r(a,b)} g(x,A,a,b,\tau)
    +
     \int_a^b g(x,A,a,\theta,\tau)
     \left(
     p(x,\tau)g(x,A,\theta,b,\tau) 
     + (1-p(x,\tau))g(x,B,\theta,b,\tau) 
     \right)d\theta
     -(b-a)g(x,A,a,b,\tau)
\end{aligned}$$
where 
$$\begin{aligned}
r(a,b)=\begin{cases}
    \min(|a|,|b|) \qquad &\text{if } ab \ge 0 \\
    0 \qquad &\text{if } 0 \in [a,b] .
\end{cases}
\end{aligned}$$
and $G_r f(x,z)$ is defined above.
On the other hand, if $a<0<b$, then
$$\begin{aligned}
\frac{d}{d\tau} 
g(x,A,a,b,\tau)
    &=
    G_{r(a,b)} g(x,A,a,b,\tau)
    +
     \int_0^b g(x,A,a,\theta,\tau)
     \left(
     p(x,\tau)g(x,A,\theta,b,\tau)
     + (1-p(x,\tau))g(x,B,\theta,b,\tau) 
     \right)d\theta \\
   &\qquad {}
    +
     \int_a^0 \left(
     p(x,\tau)g(x,A,a,\theta,\tau)
     + (1-p(x,\tau))g(x,B,a,\theta,\tau) 
     \right) g(x,A,\theta,b,\tau) d\theta  \\
   &\qquad {}
     -(b-a)g(x,A,a,b,\tau)
\end{aligned}$$

Writing $g_A$ and $g_B$ for the moment, and dropping $x$ and $\tau$,
the generator of the recombination part is, 
where $x\wedge y=\max(x,y)$ and $x \vee y = \min(x,y)$,
$$\begin{aligned}
Ig_A(a,b)
    &= 
        p \int_{a\wedge 0}^b g_A(a,\theta) g_A(\theta,b) d\theta
        + (1-p) \int_{a\wedge 0}^b g_A(a,\theta) g_B(\theta,b) d\theta
        + p \int_{a}^{b \vee 0} g_A(a,\theta) g_A(\theta,b) d\theta
        + (1-p) \int_{a\wedge 0}^b g_B(a,\theta) g_A(\theta,b) d\theta 
       \\ &\qquad {}
         -(b-a)g_A(a,b) \\
    &= 
    \int_a^b
        p \one_{\theta>0} g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \one_{\theta>0} g_A(a,\theta) g_B(\theta,b) 
        + p \one_{\theta\le0} g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \one_{\theta\le0} g_B(a,\theta) g_A(\theta,b) d\theta
       \\ &\qquad {}
         -(b-a)g_A(a,b) \\
    &= 
    \int_a^b
    \left\{
        p g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \left( 
            \one_{\theta>0} g_A(a,\theta) g_B(\theta,b) 
            + \one_{\theta\le0} g_B(a,\theta) g_A(\theta,b) 
        \right) 
        -g_A(a,b) \right\}
        d\theta   \\
    &= 
    p \int_a^b
        \left( g_A(a,\theta) g_A(\theta,b)  - g_A(a,b) \right) d\theta 
    \\ &\qquad {}
    + (1-p) \int_a^b \one_{\theta>0} \left( g_A(a,\theta) g_B(\theta,b) - g_A(a,b) \right) d\theta
    \\ &\qquad {}
    + (1-p) \int_a^b \one_{\theta\le0} \left( g_B(a,\theta) g_A(\theta,b) - g_A(a,b) \right) d\theta  .
\end{aligned}$$
$Ig_B$ is the same after exchanging $A$ with $B$, and $p$ with $1-p$.

@@

*Note:* We also know that
$$\begin{aligned}
    g(x,B,a,b,\tau) &= 0 \qquad \text{if } 0 \in [a,b] \\
\end{aligned}$$
but this will have to be arranged by the cline function.

@@

Note that if we think of $u(a,b)$ as a kernel, 
then the map $u(a,b) \mapsto \int_a^b u(a,\theta) u(\theta,b) d\theta$
is similar to the product of the kernel to itself;
on discretization this becomes like a matrix power.
This is implemented in some C code.
<!--
```{r solve_branching}
source("unifprod.R")  # for the C part
recomb_generator <- function (x,eps) {
    # the discretization of the map y(a,b) -> \int_a^b (h(a,u)*h(u,b)-h(a,b)) du
    # x is the upper triangular part of a square matrix
    #   eps is the grid spacing between locations
    x2 <- eps*unifprod_ut(x)
}
xgrid <- setup.grid.1D(x.up=-20, x.down=20, N=10)
agrid <- setup.grid.1D(x.up=0.0, x.down=0.5, N=5)
# function to find a,b coordinates from the upper triangular coordinates
ab.ord <- matrix( 0, nrow=agrid$N, ncol=agrid$N )
avals <- agrid$x.mid[row(ab.ord)[upper.tri(ab.ord,diag=TRUE)]]
bvals <- agrid$x.mid[col(ab.ord)[upper.tri(ab.ord,diag=TRUE)]]
rvals <- ifelse( avals*bvals<0, 0, pmin(avals,bvals) )
ab.coord <- function (ab) { cbind( a=avals[ab], b=bvals[ab] ) }
# we'll save the solution as an array,
#  with dimensions (x,ab)
# Here ab is upper-triangular ordered

s <- 0.1
tt <- seq(0,3,length.out=5)
fwds.soln <- forwards_pde(s=s,times=tt,grid=xgrid)

soln.array <- matrix(0.0,nrow=2*xgrid$N,ncol=agrid$N*(agrid$N+1)/2)

    sigma=1
    yinit=rep( c( rep(1.0,xgrid$N), rep(0.0,xgrid$N) ), ncol(soln.array) )

    eps <- 1e-16
    rev.pde.fn <- function (t,y,parms,...) {
        # need to apply tran.1D to columns
        # and recomb_generator to rows
        dim(y) <- dim(soln.array)
        p <- cline_interp(t,soln=fwds.soln) # freq of B
        log.p <- log(pmax(p,min(eps,min(p[p>0]))))
        log.1mp <- log(pmax(1-p,min(eps,min((1-p)[p<1]))))
        diffusion <- sapply( 1:ncol(y), function (kcol) {
                r <- rvals[kcol]
                yA <- y[1:xgrid$N,kcol]
                yB <- y[xgrid$N+(1:xgrid$N),kcol]
                tran.A <- tran.1D(C=yA, A=log.p, D=sigma^2/2, dx=xgrid, flux.up=0, flux.down=0, log.A=TRUE)$dC
                tran.B <- tran.1D(C=yB, A=log.1mp, D=sigma^2/2, dx=xgrid, flux.up=0, flux.down=0, log.A=TRUE)$dC
                c( 
                        tran.A + r * (1-p) * (yB-yA),
                        tran.B + r * p * (yA-yB)
                        )
             } )
        recombination <- apply( y, 1, recomb_generator, p=p )
        return( list( diffusion + recombination ) )
    }
    rev.soln <- ode.1D( y=yinit, times=times, func=rev.pde.fn, nspec=2*ncol(soln.array), tcrit=max(fwds.soln[,1]) )

@@

-->
